Type theories as quotient inductive-recursive types

### 2025-02-19 tex/main.tex
polish introduction (saves some space, but no effect at the end yet)

\title{Type theories as quotient inductive-recursive types}
\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.}
\author{Fredrik Nordvall Forsberg}{Department of Computer and Information Sciences, University of Strathclyde, UK \and \url{https://fredriknf.com}}{fredrik.nordvall-forsberg@strath.ac.uk}{https://orcid.org/0000-0001-6157-9288}{Supported by the Engineering and Physical Sciences Research Council [EP/Y000455/1].}
\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
\authorrunning{L.-T.~Chen, F.~Nordvall Forsberg, and T.-C.~Tsai}
\Copyright{Liang-Ting Chen, Fredrik Nordvall Forsberg, and Tzu-Chun Tsai}
\ccsdesc[500]{Theory of computation~Type theory}
\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families}
\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

\supplement{The formal development is hosted at the \textsc{GitHub} repository: \url{https://github.com/L-TChen/TTasQIIRT}}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional

@@

Metaprogramming and mechanised metatheory in a type-theoretic proof assistant relies on having an internal representation of type theory within type theory itself.
%
In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas, etc.
%
Doing this effectively seems to require several sophisticated features of the host type theory, most notably advanced inductive definitions.
%
For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed $\lambda$-calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
%
This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerably harder to work with, most likely exhausting the resources of both the proof assistant and its user.

Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types}~\cite{Altenkirch2018}, the internal representation of well typed terms could be significantly simplified, since equations such as $\beta$- and $\eta$-equality can be represented by the identity type thanks to equality constructors.
%
Yet, there is a general phenomenon, called `transport hell', when working with quotient inductive-inductive types: to make use of equalities in later constructors and when using the eliminator for the type, the equalities need to be transported across.
%

@@

In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s, for short).
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
%

@@

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
Nevertheless, \Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

\subparagraph*{Contributions}
This paper explores the use of QIIRTs and definitions by overlapping patterns to define type theories in intensional type theory.
In particular, we give
\begin{itemize}
  \item definitions of parallel substitution calculus, type theories with a universe and $\Pi$-types,
  \item with their elimination principles and the standard model;
  \item a comparison with these definitions using quotient inductive-inductive types.
\end{itemize}

\subparagraph*{Plan of the paper}
In the rest of the paper, we introduce the metatheory in \cref{sec:meta}.
Then, we develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, followed by their elimination principles and usage in \cref{sec:models}.
We outline the \Agda formal implementation in \cref{sec:formal-implementation}, and we conclude with related and future work in \cref{sec:conclusion}.

@@

Going back and forth from one definition to the other gives us the equivalences
\begin{restatable}{theorem}{equivalence}
  Type theory defined as a QIIRT and a QIIT are equivalent, i.e.\ 
  \begin{itemize}
    \item $\eqv{\Ctx_{\IR}}{\Ctx_{\I}}$, 
    \item $\eqv{\left((\Gamma : \Ctx_{\IR})\to\Ty_{\IR}\;\Gamma\right)}{\left((\Gamma : \Ctx_{\I})\to\Ty_{\I}\;\Gamma\right)}$,
    \item $\eqv{\left((\Gamma\;\Delta : \Ctx_{\IR}) \to \Sub_{\IR}\;\Gamma\;\Delta\right)}%
               {\left((\Gamma\;\Delta: \Ctx_{\I}) \to \Sub_{\I}\;\Gamma\;\Delta\right)}$, and 
    \item $\eqv{\left((\Gamma : \Ctx_{\IR}) \to (A : \Ty_{\IR}\;\Gamma) \to \Tm_{\IR}\;\Gamma\;A\right)}
               {\left((\Gamma : \Ctx_{\I}) \to (A : \Ty_{\I}\;\Gamma) \to \Tm_{\I}\;\Gamma\;A\right)}$.
  \end{itemize}
\end{restatable}

>>>>

\title{Type theories as quotient inductive-recursive types}
\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{The National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.}
\author{Fredrik Nordvall Forsberg}{Department of Computer and Information Sciences, University of Strathclyde, UK \and \url{https://fredriknf.com}}{fredrik.nordvall-forsberg@strath.ac.uk}{https://orcid.org/0000-0001-6157-9288}{The Engineering and Physical Sciences Research Council [EP/Y000455/1].}
\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{The National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
\authorrunning{L.-T.~Chen, F.~Nordvall Forsberg, and T.-C.~Tsai}
\Copyright{Liang-Ting Chen, Fredrik Nordvall Forsberg, and Tzu-Chun Tsai}
\ccsdesc[500]{Theory of computation~Type theory}
\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families, type theory in type theory.}
\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}./}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional

@@

Metaprogramming and mechanised metatheory in a type-theoretic proof assistant relies on having an internal representation of type theory within type theory itself.
%
In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas.
%
Doing this effectively seems to require sophisticated features of the host type theory, most notably advanced inductive definitions.
%
For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed $\lambda$-calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
%
This is not just a matter of pure strength, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerably harder to work with, most likely exhausting the resources of both the proof assistant and its user.

Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types} (QIITs)~\cite{Altenkirch2018}, the internal representation of well typed terms could be significantly simplified, since equality constructors can be used to represent equations such as $\beta$- and $\eta$-equality.
%
Yet, there is a general phenomenon, called `transport hell', when working with QIITs: to make use of equalities in later constructors and when using eliminators, equalities need to be transported across,
leading to intricate higher-dimensional equations between equations.

@@

In this paper, we show that the development can be streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s).
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and precursors by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: pattern matching definitions with overlapping patterns.
%
This is helpful to avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees~\cite{Cockx2017}.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with QIITs alone).
%

@@

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are, so far, not yet well understood or foundationally well motivated.
Nevertheless, the proof assistant \Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

\subparagraph*{Contributions}
This paper explores the use of QIIRTs and definitions by overlapping patterns to define type theories in intensional type theory. Our contributions are:
\begin{itemize}
  \item We give definitions of the parallel substitution calculus and type theories with a universe and $\Pi$-types as a QIIRT (\cref{sec:QIIRTs}).
  \item We show how elimination principles for reasoning about the type theory can be derived, and use it to construct its standard model (\cref{sec:models}).
  \item We compare our definition to definitions using quotient inductive-inductive types, and prove that the two definitions are actually equivalent (\cref{thm:equivalence}).
\end{itemize}

%\subparagraph*{Plan of the paper}
%In the rest of the paper, we introduce the metatheory in \cref{sec:meta}.
%Then, we develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, follo%wed by their elimination principles and usage in \cref{sec:models}.
%We outline the \Agda formal implementation in \cref{sec:formal-implementation}, and we conclude with related and future work in \cref{sec:conclusion}.

@@

Going back and forth from one definition to the other gives us the equivalences
\begin{restatable}{theorem}{equivalence}\label{thm:equivalence}
  Type theory defined as a QIIRT and a QIIT are equivalent, i.e.\ 
  \begin{itemize}
    \item $\eqv{\Ctx_{\IR}}{\Ctx_{\I}}$, 
    \item $\eqv{\left((\Gamma : \Ctx_{\IR})\to\Ty_{\IR}\;\Gamma\right)}{\left((\Gamma : \Ctx_{\I})\to\Ty_{\I}\;\Gamma\right)}$,
    \item $\eqv{\left((\Gamma\;\Delta : \Ctx_{\IR}) \to \Sub_{\IR}\;\Gamma\;\Delta\right)}%
               {\left((\Gamma\;\Delta: \Ctx_{\I}) \to \Sub_{\I}\;\Gamma\;\Delta\right)}$, and 
    \item $\eqv{\left((\Gamma : \Ctx_{\IR}) \to (A : \Ty_{\IR}\;\Gamma) \to \Tm_{\IR}\;\Gamma\;A\right)}
               {\left((\Gamma : \Ctx_{\I}) \to (A : \Ty_{\I}\;\Gamma) \to \Tm_{\I}\;\Gamma\;A\right)}$.
  \end{itemize}
\end{restatable}
